DaLi: Database as a Library

### 2017-02-20 paper/motivation.tex
Persistence

Another downside with effect-generating replicated data type libraries
is that they don't compose. For instance, let us say that an
application wants to use two replicated counters, $c_1$ and $c_2$,
bound by an invariant $c_2 \ge c_1$, which the application duly the
invariant when updating $c_2$ and $c_1$. Nonetheless, it may still
witness anamolous states that violate the invariant because updates to
$c_1$ and $c_2$ are applied independently in any order. For example,
if a replica increments both the counters by 1 each, the \C{read} at
another replica, where $(c_1,c_2) = (2,2)$ may witness the increment
to $c_1$, but not $c_2$, thus violating the invariant. The intention
to atomically apply both the effects cannot be captured in the absence
of external mechanisms to compose effects (such as effect-based
transactions).

>>>>

Another downside with effect-generating replicated data type libraries
is that they don't compose. For instance, let us say that an
application wants to use two replicated counters, $c_1$ and $c_2$,
bound by an invariant $c_2 \ge c_1$, which the application duly
enforces when updating $c_2$ and $c_1$. Nonetheless, it may still
witness anamolous states that violate the invariant because updates to
$c_1$ and $c_2$ are applied independently in any order. For example,
if a replica increments both the counters by 1 each, the \C{read} at
another replica, where $(c_1,c_2) = (2,2)$ may witness the increment
to $c_1$, but not $c_2$, thus violating the invariant. The intention
to atomically apply both the effects cannot be captured in the absence
of external mechanisms to compose effects (such as effect-based
transactions~\cite{pldi15}).

Persistence adds another dimension to the problem. Functional data
structures are often large in-memory linked structures that admit
functional updates by creating newer versions that share most of their
structure with previous versions. Sharing is the key to efficiency,
without which every functional update takes time that is at least
linear in the size of the data structure. Unfortunately, the
straightforward way to persist a data structure on disk in a
machine-agnostic format (e.g., JSON) requires serialization, which is
linear in the size of the data structure. An alternative is to
maintain a separate on-disk copy of the data and administer it via a
database system. The downside is that the programmer is now required
to reason in terms of a lower-level abstraction (e.g., a relational
data model) in order to maintain coherence between in-memory and
on-disk representations. Safety guarantees over the in-memory data do
not carry over to the disk, which introduces additional complications
and the possibility of new bugs. Thus, adding persistence to
applications while preserving high-level safety gurantees offered by
the language abstraction is a non-trivial problem.
